---
title: "Directed Acyclic Graphs (DAGs)"
subtitle: "EC 607 Metrics, Tutorial 7"
author: "Philip Economides"
date: "Spring 2021"
output:
  xaringan::moon_reader:
    css: ['default', 'metropolis', 'metropolis-fonts', 'my-css.css']
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---
class: middle

```{R setup, include = F}
# devtools::install_github("dill/emoGG")
library(pacman)
p_load(
  broom, tidyverse, pracma, tstools,
  latex2exp, ggplot2, ggthemes, ggforce, viridis, extrafont, gridExtra,
  kableExtra, snakecase, janitor,
  data.table, dplyr, estimatr,
  lubridate, knitr, parallel,
  lfe,scales,ggdag,
  here, magrittr, kableExtra, snakecase, janitor, lubridate,
  data.table, knitr, jtools, huxtable, estimatr, haven, ipumsr
)
# Define pink color
red_pink <- "#e64173"
turquoise <- "#20B2AA"
orange <- "#FFA500"
red <- "#fb6107"
blue <- "#2b59c3"
green <- "#8bb174"
grey_light <- "grey70"
grey_mid <- "grey50"
grey_dark <- "grey20"
purple <- "#6A5ACD"
slate <- "#314f4f"
# Dark slate grey: #314f4f
# Knitr options
opts_chunk$set(
  comment = "#>",
  fig.align = "center",
  fig.height = 7,
  fig.width = 10.5,
  warning = F,
  message = F
)
opts_chunk$set(dev = "svg")
options(device = function(file, width, height) {
  svg(tempfile(), width = width, height = height)
})
options(crayon.enabled = F)
options(knitr.table.format = "html")
```

```{css, echo = F, eval = F}
@media print {
  .has-continuation {
    display: block !important;
  }
}
```


## Today

- Recap DAGs
- Simple DAG
- Adding Chains
- Collider Bias

---

layout: true

# Recap
---

<br>

*  Graphical representation of a chain of causal effects. 

--

* Explain causality in terms of counterfactuals. 

--

* Nodes for random variables, random variables assumed to follow some data-generating process.

--

* Causal effects defined as comparison between two states of world;<br> one state when intervention occured on some value and another state where it did not occur under some other intervention. 

--

* Effects are eiether direct or mediated by a third variable.  


---

layout: true

# Simple DAG

---

```{r}
p_load(dagitty,ggdag)

g <- dagitty("dag{ 
  a -> b ;
  b -> c ;
  d -> c
 }")
#plot( graphLayout(g))
```


```{r}
# Evaluating linkages
parents(g, "c")
children(g, "b")
```

---

### Simple DAG



```{r, echo=FALSE, out.width="90%"}
p_load(dagitty, ggdag)

#dagitty(x, layout = FALSE)
g <- dagitty("dag{ 
  a -> b ;
  b -> c ;
  d -> c
 }")
plot( graphLayout(g))
```



---

First, assign your random variable linkages and their coordinates.

```{r}
p_load(data.table)

dag_full = dagify(
  Y ~ D,
  Y ~ W,
  D ~ W,
  coords = tibble(
    name = c("Y", "D", "W"),
    x = c(1, 3, 2),
    y = c(2, 2, 1)  ) )
```


---

We can include information for segments in `data.table` format.

```{r}
# Convert to data.table
dag_dt = dag_full %>% ggplot2::fortify() %>% setDT()
dag_dt[, `:=`(
  path1 = (name == "D" & to == "Y") | (name == "Y"),
  path2 = (name == "D" & to == "W") | (name == "W" & to == "Y") | (name == "Y")
)]
```

---

Having set up our mapping, we just need co-ordinates for causal arrows.

```{r}
# Shorten segments
mult = 0.15
dag_dt[, `:=`(
  xa = x + (xend-x) * (mult),
  ya = y + (yend-y) * (mult),
  xb = x + (xend-x) * (1-mult),
  yb = y + (yend-y) * (1-mult)
)]
```


---

<br>

Using ggplot and the segment points, we can begin visualizing. 


```{r}
# Plot the full DAG
p1 = ggplot(
  data = dag_dt,
  aes(x = x, y = y, xend = xend, yend = yend)
) +
geom_point(
  size = 20,
  fill = "white",
  color = slate,
  shape = 21,
  stroke = 0.6
) 
```

---

```{r, echo=FALSE}
p1
```

---

<br>

We then add our causal arrows using segment info.

```{r, warning=FALSE, message=FALSE}
p2 = p1+
geom_curve(
  aes(x = xa, y = ya, xend = xb, yend = yb),
  curvature = 0,
  arrow = arrow(length = unit(0.07, "npc")),
  color = slate,
  size = 1.2,
  lineend = "round"
) 
```

---

```{r, echo=FALSE}
p2
```


---

Lastly, update the nodes with corresponding variables and adjust appearance. 

```{r}
p3 = p2+
geom_text(
  data = dag_dt[,.(name,x,y,xend=x,yend=y)] %>% unique(),
  aes(x = x, y = y, label = name),
  family = "Fira Sans Medium",
  size = 8,
  color = slate,
  fontface = "bold"
) +
theme_void() +
theme(
  legend.position = "none",
) +
coord_cartesian(
  xlim = c(dag_dt[,min(x)]*0.95, dag_dt[,max(x)]*1.05),
  ylim = c(dag_dt[,min(y)]*0.8, dag_dt[,max(y)]*1.1)
)
```

---

```{r, echo=FALSE}
p3
```

---

layout: true

# Chains

---

Let us denote *associated* changes of A on C through B using a DAG. 

```{r}
bb3_ex = dagify(
  B ~ A, C ~ B,
  coords = tibble(
    name = LETTERS[1:3],
    x = -1:1, y = 0  )
)
# Convert to data.table
bb3_dt = bb3_ex %>% fortify() %>% setDT()
bb3_dt
```

---

Again we'll plot our coordinates for the segments. 

```{r}
# Shorten segments
mult = 0.25
bb3_dt[, `:=`(
  xa = x + (xend-x) * (mult),
  ya = y + (yend-y) * (mult),
  xb = x + (xend-x) * (1-mult),
  yb = y + (yend-y) * (1-mult)
)]
```

---

We can begin visualizing. We'll start simple with causal arrows.

```{r}
# Plot the DAG
gg_chain = ggplot(
  data = bb3_dt,
  aes(x = x, y = y)
) +
geom_curve(
  aes(x = xa, y = ya, xend = xb, yend = yb),
  curvature = 0,
  arrow = arrow(length = unit(0.09, "npc")),
  color = purple,
  size = 0.9,
  lineend = "round"
)
```

---

<br>

```{r, out.width="80%", echo=FALSE}
gg_chain
```

---

Adding our nodes.

```{r}
gg_chain = gg_chain +
geom_point(
  size = 20,
  fill = "white",
  color = purple,
  shape = 21,
  stroke = 0.6
)
```



---

<br>

```{r, out.width="80%", echo=FALSE}
gg_chain
```

---

Node text and cleaning up the appearance. 

```{r}
gg_chain = gg_chain   +
geom_text(
  aes(x = x, y = y, label = name),
  family = "Fira Sans Medium",
  size = 8,
  color = purple,
  fontface = "bold"
) +
theme_void() +
theme(
  legend.position = "none",
) +
coord_cartesian(
  xlim = c(-1.5, 1.5),
  ylim = c(-1, 0.5)
)
```

---

<br>

```{r, out.width="80%", echo=FALSE}
gg_chain
```

---

<br>

The chain is easily added to this chart. 

```{r, out.width="80%", eval=FALSE}
curve_dt = tibble(
  x = c(-1, 0, 1),
  y = c(0, -0.8, 0)
) %>% spline(n = 101) %>% as.data.table()

gg_chain +
geom_line(
  data = curve_dt,
  color = orange,
  linetype = "dashed",
  size = 0.8
)
```


---

<br>

```{r, out.width="80%", echo=FALSE}
curve_dt = tibble(
  x = c(-1, 0, 1),
  y = c(0, -0.8, 0)
) %>% spline(n = 101) %>% as.data.table()

gg_chain +
geom_line(
  data = curve_dt,
  color = orange,
  linetype = "dashed",
  size = 0.8
)
```



Notice anything off? 

---

layout: true

# Collider Bias

---

```{r, message=FALSE, warning=FALSE, eval=FALSE}
p_load(tidyverse, jtools)

tb <- tibble(
  female = ifelse(runif(10000)>=0.5,1,0),
  ability = rnorm(10000),
  discrimination = female,
  occupation = 1 + 2*ability + 0*female - 2*discrimination + rnorm(10000),
  wage = 1 - 1*discrimination + 1*occupation + 2*ability + rnorm(10000) 
)
```

How would we represent these random variables in a DAG?

--

$$
\begin{aligned}
&F \implies D, \ \ D \implies O, \ \  D \implies Y\\
&A \implies O, \ \ A \implies Y, \ \  O \implies Y
\end{aligned}
$$

--

Using an example from [Causal Inference: The Mixtape](https://mixtape.scunning.com/dag#more-examples-of-collider-bias)

---

$$
\begin{aligned}
&F \implies D, \ \ D \implies O, \ \  D \implies Y\\
&A \implies O, \ \ A \implies Y, \ \  O \implies Y
\end{aligned}
$$

--

```{r, eval=FALSE}
# F, O, D, Y , A   (X range: 0,3 , Y range: 0,5)
bb3_ex = dagify(
  D ~ F, O ~ D, Y ~ D, 
  O ~ A, Y ~ A, Y ~ O,
  coords = tibble(
    name = c("F","O", "D", "Y", "A"),
    x =  c(0, 0, 2, 3, 3),
    y = c(3, 0, 5, 3, 0)
  )
)
# Convert to data.table
bb3_dt = bb3_ex %>% fortify() %>% setDT()
# Shorten segments
mult = 0.15
bb3_dt[, `:=`(
  xa = x + (xend-x) * (mult),
  ya = y + (yend-y) * (mult),
  xb = x + (xend-x) * (1-mult),
  yb = y + (yend-y) * (1-mult)
)]
```

---

$$
\begin{aligned}
&F \implies D, \ \ D \implies O, \ \  D \implies Y\\
&A \implies O, \ \ A \implies Y, \ \  O \implies Y
\end{aligned}
$$


```{r, echo=FALSE}
# F, O, D, Y , A   (X range: 0,3 , Y range: 0,5)
bb3_ex = dagify(
  D ~ F,
  O ~ D,
  Y ~ D, 
  O ~ A,
  Y ~ A,
  Y ~ O,
  coords = tibble(
    name = c("F","O", "D", "Y", "A"),
    x =  c(0, 0, 2, 3, 3),
    y = c(3, 0, 5, 3, 0)
  )
)
# Convert to data.table
bb3_dt = bb3_ex %>% fortify() %>% setDT()
# Shorten segments
mult = 0.15
bb3_dt[, `:=`(
  xa = x + (xend-x) * (mult),
  ya = y + (yend-y) * (mult),
  xb = x + (xend-x) * (1-mult),
  yb = y + (yend-y) * (1-mult)
)]
```



---

$$
\begin{aligned}
&F \implies D, \ \ D \implies O, \ \  D \implies Y\\
&A \implies O, \ \ A \implies Y, \ \  O \implies Y
\end{aligned}
$$

```{r, out.width="80%", echo=FALSE}
gg_discr = ggplot(
  data = bb3_dt,
  aes(x = x, y = y)
) +
  geom_curve(
  aes(x = xa, y = ya, xend = xb, yend = yb),
  curvature = 0,
  arrow = arrow(length = unit(0.03, "npc")),
  color = purple,
  size = 0.9,
  lineend = "round"
) +
geom_point(
  size = 20,
  fill = "white",
  color = purple,
  shape = 21,
  stroke = 0.6
)+
geom_text(
  aes(x = x, y = y, label = name),
  family = "Fira Sans Medium",
  size = 8,
  color = purple,
  fontface = "bold"
) + theme_void() +
theme(
  legend.position = "none",
) +
coord_cartesian(
  xlim = c(-0.5, 3.5),
  ylim = c(-0.5, 5.5)
)
# Plot it
gg_discr
```

---

```{r, message=FALSE, warning=FALSE, echo=FALSE}
p_load(tidyverse, jtools)

tb <- tibble(
  female = ifelse(runif(10000)>=0.5,1,0),
  ability = rnorm(10000),
  discrimination = female,
  occupation = 1 + 2*ability + 0*female - 2*discrimination + rnorm(10000),
  wage = 1 - 1*discrimination + 1*occupation + 2*ability + rnorm(10000) 
)

lm_1 <- lm(wage ~ female, tb)
lm_2 <- lm(wage ~ female + occupation, tb)
lm_3 <- lm(wage ~ female + occupation + ability, tb)

export_summs(lm_1,lm_2,lm_3, model.names= c("Biased Unconditional", 
                                            "Biased",
                                            "Unbiased Conditional"))
```

---

exclude: true

```{R generate pdfs, include = F, eval = F}
#remotes::install_github('rstudio/pagedown')
library(pagedown)
pagedown::chrome_print("Tutorial-7.html", output = "Tutorial-7.pdf")
#pagedown::chrome_print("Tutorial Slides 1-nopause.html", output = "01-research-r-nopause.pdf")
```