---
title: "Regression Analysis"
subtitle: "EC 607 Metrics, Tutorial 6"
author: "Philip Economides"
date: "Spring 2021"
output:
  xaringan::moon_reader:
    css: ['default', 'metropolis', 'metropolis-fonts', 'my-css.css']
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---
class: middle

```{R setup, include = F}
# devtools::install_github("dill/emoGG")
library(pacman)
p_load(
  broom, tidyverse, pracma, tstools,
  latex2exp, ggplot2, ggthemes, ggforce, viridis, extrafont, gridExtra,
  kableExtra, snakecase, janitor,
  data.table, dplyr, estimatr,
  lubridate, knitr, parallel,
  lfe,scales,
  here, magrittr, kableExtra, snakecase, janitor, lubridate,
  data.table, knitr, jtools, huxtable, estimatr, haven, ipumsr
)
# Define pink color
red_pink <- "#e64173"
turquoise <- "#20B2AA"
orange <- "#FFA500"
red <- "#fb6107"
blue <- "#2b59c3"
green <- "#8bb174"
grey_light <- "grey70"
grey_mid <- "grey50"
grey_dark <- "grey20"
purple <- "#6A5ACD"
slate <- "#314f4f"
# Dark slate grey: #314f4f
# Knitr options
opts_chunk$set(
  comment = "#>",
  fig.align = "center",
  fig.height = 7,
  fig.width = 10.5,
  warning = F,
  message = F
)
opts_chunk$set(dev = "svg")
options(device = function(file, width, height) {
  svg(tempfile(), width = width, height = height)
})
options(crayon.enabled = F)
options(knitr.table.format = "html")
```

```{css, echo = F, eval = F}
@media print {
  .has-continuation {
    display: block !important;
  }
}
```


## Today

- Ordinary Least Squares
- Regression Output
- Fixed-Effect Models
- Instrumental Variables
- Standard Errors 

---

layout: true

# Regression Analysis
---

<br>

With such a wide array of package options, confusion can easily arise. 

--

Additionally, some packages are more cumbersome than others.

--

In order to alleviate some of these concerns, I'll run through a set of recommended packages. 


---

layout: true

# Ordinary Least Squares

---

<br>

While `lm()` from the stats package is a common go to for fitting linear models, the `fixest` package is an alternative worth considering. 

--

```{r}
# Lets generate some data to fit into our OLS functions
set.seed(12345)
X = runif(10000, 2, 5)
e = rnorm(10000, 0, 1)
Y = 4 + 5*X + e

# Do we return correct coefficients?
Xm = matrix(c(rep(1,10000), X), ncol=2)
t(solve(t(Xm) %*% Xm) %*% (t(Xm) %*% Y))
```

---

<br>

.pull-left[
```{r}
d_frame = bind_cols(Y,X) %>% 
  rename(Y=...1, X=...2)
a = lm(Y ~ X, data=d_frame)
a %>% tidy() %>% select(1:3)
```
]

--

.pull-right[
```{r}
# Alternative means
p_load(fixest)
b=feols(Y ~ X, data=d_frame)
b %>% tidy()%>% select(1:3)
```
]

---

layout: true

# Regression Output

---

```{r}
# Using export_summs()
p_load(jtools)
export_summs(a,b, model.names = c("lm()", "feols()"), 
    coefs = c("constant" = "(Intercept)", "covariate" = "X"))
```

---

```{r}
# Using stargazer, b object unrecognized. type=text,latex,html
p_load(stargazer)
stargazer(a, title="Regression Results", type="html",
          single.row=TRUE,ci=TRUE, ci.level=0.95,
          omit.stat=c("f", "ser"))
```

---

<table style="text-align:center"><caption><strong>Regression Results</strong></caption>
<tr><td colspan="2" style="border-bottom: 1px solid black"></td></tr><tr><td style="text-align:left"></td><td><em>Dependent variable:</em></td></tr>
<tr><td></td><td colspan="1" style="border-bottom: 1px solid black"></td></tr>
<tr><td style="text-align:left"></td><td>Y</td></tr>
<tr><td colspan="2" style="border-bottom: 1px solid black"></td></tr><tr><td style="text-align:left">X</td><td>5.010<sup>***</sup> (4.987, 5.033)</td></tr>
<tr><td style="text-align:left">Constant</td><td>3.966<sup>***</sup> (3.884, 4.048)</td></tr>
<tr><td colspan="2" style="border-bottom: 1px solid black"></td></tr><tr><td style="text-align:left">Observations</td><td>10,000</td></tr>
<tr><td style="text-align:left">R<sup>2</sup></td><td>0.949</td></tr>
<tr><td style="text-align:left">Adjusted R<sup>2</sup></td><td>0.949</td></tr>
<tr><td colspan="2" style="border-bottom: 1px solid black"></td></tr><tr><td style="text-align:left"><em>Note:</em></td><td style="text-align:right"><sup>*</sup>p<0.1; <sup>**</sup>p<0.05; <sup>***</sup>p<0.01</td></tr>
</table>

---



```{r}
# Using kable()
a %>% tidy() %>% kable()
b %>% tidy() %>% kable()
```

---

```{r}
b %>% tidy() %>% 
  mutate(
    p.value = scales::pvalue(p.value),
    term = c("Intercept", "Approval rate")
  ) %>%
  kable(caption = "Estimates using Simulated Data",
    col.names = c("Predictor", "B", "SE", "t", "p"),
    digits = c(0, 2, 3, 2, 3),
    align = c("l", "r", "r", "r", "r")
    )%>%
  kable_classic(full_width = F, html_font = "Cambria")
```

Check out the kableExtra [vignette](https://cran.r-project.org/web/packages/kableExtra/vignettes/awesome_table_in_html.html) for more awesome tables in HTML. 

---

layout: true

# Fixed-Effect Models

---

Using Laurent Berge's [FE introduction](https://cran.r-project.org/web/packages/fixest/vignettes/fixest_walkthrough.html)

```{r}
p_load(fixest)
# Intra-EU15 trade
data(trade) 
glimpse(trade)
```

---


### Structure

```{r, eval=FALSE}
#Estimates OLS with any number of fixed-effects.
feols( dep_var ~ cov1 + cov2 | 
         fe_1 + fe_2, data=df_name)
# Note:  fe_1^fe_2 to create year-industry FE. 
# Note: Can specify clustered se in summary.

#Estimates GLM models with any number of fixed-effects.
feglm( dep_var ~ cov1 + cov2 | 
         fe_1 + fe_2, data=df_name)
# Note:  Default is a poisson model.

```


Estimation is performed using parallel computing which you can control using the argument `nthreads`
 
---

### Example

```{r}
gravity_pois = fepois(Euros ~ log(dist_km) | 
          Origin + Destination + Product + Year, data=trade)
# Check out the details that stem from raw output. Very informative!
print(gravity_pois)
```


---

### Clustering in FE models

```{r}
# Individual clustering
a=summary(gravity_pois, cluster="Origin") %>% tidy()
b=summary(gravity_pois, cluster="Destination") %>% tidy()
c=summary(gravity_pois, cluster="Product") %>% tidy()
d=summary(gravity_pois, cluster="Year") %>% tidy()

# Multi-way clustering
e=summary(gravity_pois, cluster=c("Origin", "Destination")) %>% tidy()
f=summary(gravity_pois, cluster=c("Product", "Year")) %>% tidy()

# Set of results
c((rbind(a,b,c,d,e,f)[,3]))
```

---

### Results

```{r, eval=FALSE}
# OLS model
gravity_ols = feols(log(Euros) ~ log(dist_km) | 
              Origin + Destination + Product + Year, trade)
# Negative Binomial
gravity_negbin = fenegbin(Euros ~ log(dist_km) | 
              Origin + Destination + Product + Year, trade)
# Present
etable(gravity_pois, gravity_negbin, gravity_ols,
      se = "twoway", #defaults to the first two FE's 
      subtitles = c("Poisson", "Negative Binomial", "Gaussian")
      # tex=TRUE will report results in LaTeX table format)
```

--

For more on [exporting etable](https://cran.r-project.org/web/packages/fixest/vignettes/exporting_tables.html) with preset style.tex including AER format. 

--

```{r}
# Multiple estimations: Stepwise FE starting with none. 
res_multi = fepois(Euros ~ log(dist_km) | 
            csw0(Year, Destination, Origin), trade)
```


---

layout: true

# Instrumental Variables

---

<br>

Two solid options here;

`fixest`: Benefits to sticking with the same package throughout! 

`ivreg`: various regression diagnostics are supported.

--

<br>

Material on [ivreg diagnostics](https://john-d-fox.github.io/ivreg/articles/Diagnostics-for-2SLS-Regression.html)

---

<br>

`fixest`: 2SLS performed by adding formula for link between the endog(s) and the IV(s).

--

```{r, eval=FALSE}
# Standard Layout
reg_out = feols(dep ~ exo_cov1 + exo_cov2 | 
                  fe1 + fe2 | # Fixed Effects
                endo1 + endo2 ~ iv1 + iv2)
```

--

1. Formula for regression using exogenous covariates
2. Set of fixed effects 
3. Formula for endogenous variables and IVs (stage one)

---

### Example

```{r}
base = iris
names(base) = c("y", "x1", "x_endo_1", "x_inst_1", "fe")
set.seed(2)
base$x_inst_2 = 0.2 * base$y + 0.2 * base$x_endo_1 + rnorm(150, sd = 0.5)
base$x_endo_2 = 0.2 * base$y - 0.2 * base$x_inst_1 + rnorm(150, sd = 0.5)

est_iv_fe = feols(y ~ x1 | fe | x_endo_1 + x_endo_2 ~ x_inst_1 + x_inst_2, base)
est_iv_fe %>% tidy()
```

---

```{r}
summary(est_iv_fe)
```

---

```{r}
summary(est_iv_fe, stage = 1)[1]
```

---

```{r}
summary(est_iv_fe, stage = 1)[2]
```

---
### Results

--

<br>

To produce a LaTeX output with associated first stage regressions:

--

```{r, eval=FALSE}
etable(summary(est_iv_fe, stage = 1:2), 
       fitstat = ~ . + ivfall + ivwaldall.p, tex=TRUE)
```

---

`ivreg`: 2SLS performed by adding formula for link between the endog(s) and the IV(s).

<br>

--

```{r, eval=FALSE}
p_load(ivreg)
# Standard Layout
reg_out = ivreg(y ~ x1 + x2 | x1 + z1 + z2)
# x1, z1, z2 are all exogenous values
# x2 is the endogenous term
```
<br>

--

summary(reg_out) will yield diagnostic tests for weakness and over-ID. 

---

layout: true

# Standard Errors 

---

<br>

Often times empirical research requires the correct choice of SEs.

--

More times than not, individuals will repeatedly run a given model under different SE settings.

--

As highlighted previously, model only needs to be run once to assess SE options. 

--

We will run a model .hi-slate[only once] and adjust SEs on the backend. 

--

For more details, check out Grant McDermott's [piece on SEs](https://grantmcdermott.com/better-way-adjust-SEs/#motivation) 

---

### Sandwich


```{r}
p_load(sandwich,lmtest)
data('PetersenCL')

m = lm(y ~ x, data = PetersenCL)
summary(m)
```

---

### Sandwich

```{r}
# specification of the covariance matrix: homoskedastic
coeftest(m, vcov = vcovHC(m, type="const"))

# More SEs
type = c("HC3", "const", "HC", "HC0", "HC1", "HC2", "HC4", "HC4m", "HC5")
```

---

### More SEs! 

```{r}
vc = list(
  "Standard"              = vcov(m),
  "Sandwich (basic)"      = sandwich(m),
  "Clustered"             = vcovCL(m, cluster = ~ firm),
  "Clustered (two-way)"   = vcovCL(m, cluster = ~ firm + year),
  "HC3"                   = vcovHC(m),
  "Andrews' kernel HAC"   = kernHAC(m),
  "Newey-West"            = NeweyWest(m),
  "Bootstrap"             = vcovBS(m),
  "Bootstrap (clustered)" = vcovBS(m, cluster = ~ firm)
  )
```

Be sure to check out the links for further details on each of these items. 


---

layout: true

# Binary Outcome Models

---

<br>

* Two possible outcomes. Normally estimated by MLE.

--

* Probability of outcome I equals $p$, probabiliy of outcome II equal $1-p$. 

--

* This $p$ will vary across individuals as a function of regressors. 

--

* Logit and probit specify two function forms for this prob as function of regressors. 

--

* OLS ignores discreteness of $y_i$, would not constrain $p \in [0,1]$ 


---

### Example: Fishing Choice Model

--

$$
\begin{aligned}
y_i =& \{1, 0\}  \text{ fishing from charter boat (1) or pier (0)}\\
x_i =& \ln(\text{relp}_i) = \ln(\text{price}_{\text{charter,i}} / \text{price}_{\text{pier,i}}) 
\end{aligned}
$$

--

$p_i$ decreases as the price of the chartered boat service increases and price of pier decreases. These prices vary across individuals. 




--

Logit Model ensures that 0 < $p_i$ < 1.

$$
\begin{aligned}
p_i = Pr[ \ y_i=1 \ | \ x_i \ ] = & \frac{\exp(\beta_1 + \beta_2 x_i)}{1+\exp(\beta_1 + \beta_2 x_i)}\\
\text{Marginal Effect: } \frac{\partial p_i}{\partial x_i} = & \frac{\exp(\beta_1 + \beta_2 x_i)}{(1+ \exp(\beta_1 + \beta_2 x_i))^2} \beta_2
\end{aligned}
$$
--

Actual marginal effect approximates to $\bar y (1- \bar y) \hat \beta_2$

```{r, include=FALSE}
(452/630)*(1- (452/630))*(-1.823)
```

---

Probit Model also ensures that 0 < $p_i$ < 1.


$$
\begin{aligned}
& p_i =Pr[ \ y_i=1 \ | \ x_i \ ] =   \ \phi (\beta_1 + \beta_2 x_i) = \int_{-\infty}^{\beta_1 + \beta_2 x_i} (2\pi)^{-\frac{1}{2}} \ e^{- \frac{z^2}{2}} dz  \\
& \text{where } \phi \text{ represents the CDF for the standard normal }\\
& \text{Marginal Effect: } \frac{\partial p_i}{\partial x_i} =   \phi (\beta_1 + \beta_2 x_i) \beta_2
\end{aligned}
$$
--

What about in the case of a vector of regressors $x$?

$p_i \equiv Pr[ y_i = 1 | x ] = F(x_i' \beta)$ where F(.) is a specified function.  

--

$F(.)$: CDF of logistic distribution (logit), standard normal CDF (probit).

--

Check out .hi-slate[Cameron & Trivedi, Microeconomic Methods and Applications] for more. 

---

How do we run this in R? `fixest::feglm()` or `stats::glm()`

```{r}
df_trade <- trade %>% mutate(eur_bin = ifelse(Euros >= 4.615e+06, 1, 0))
res_trade <- feglm(eur_bin ~ log(dist_km) | 
        Origin + Destination + Product + Year, df_trade, family=binomial)
res_trade2 <- feols(eur_bin ~ log(dist_km) | 
        Origin + Destination + Product + Year, df_trade)
export_summs(res_trade, res_trade2,
             model.names=c("Logit", "OLS"),
             statistics=c("nobs", "r.squared","pseudo.r.squared"))
```


---

Propensity Scores?

```{r, echo=FALSE, out.width="80%"}
#plot(res_trade[["fitted.values"]])
phat_l = c(res_trade[["fitted.values"]])
phat_o = c(res_trade2[["fitted.values"]])
plot_df = bind_cols(df_trade[,5], phat_l, phat_o, df_trade[,7])
colnames(plot_df) <- c("dist", "phat_l", "phat_o", "eur_bin")
plot_df <- plot_df %>% arrange(desc(phat_l))
plot_dfo <- plot_df %>% arrange(desc(phat_o))

ggplot(plot_df, aes(x=1:nrow(plot_df), y=phat_l))+
  geom_point( color="blue") + xlab(NULL) + ylab("p")+
  geom_line(data=plot_dfo, aes(x=1:nrow(plot_df), y=phat_o), linetype = "longdash")+
  geom_point(aes(x=1:nrow(plot_df), y= eur_bin), alpha=0.5) + theme_pander() 
```


---

exclude: true

```{R generate pdfs, include = F, eval = F}
#remotes::install_github('rstudio/pagedown')
library(pagedown)
pagedown::chrome_print("Tutorial-6.html", output = "Tutorial-6.pdf")
#pagedown::chrome_print("Tutorial Slides 1-nopause.html", output = "01-research-r-nopause.pdf")
```