---
title: "Loops and Functions"
subtitle: "EC 607 Metrics, Tutorial 4"
author: "Philip Economides"
date: "Spring 2021"
output:
  xaringan::moon_reader:
    css: ['default', 'metropolis', 'metropolis-fonts', 'my-css.css']
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---
class: middle

```{r CSS, eval=FALSE, include=FALSE}
#css: ['default', 'metropolis', 'metropolis-fonts', 'my-css.css']
```

```{R setup, include = F}
# devtools::install_github("dill/emoGG")
library(pacman)
p_load(
  broom, tidyverse, pracma, tstools,
  latex2exp, ggplot2, ggthemes, ggforce, viridis, extrafont, gridExtra,
  kableExtra, snakecase, janitor,
  data.table, dplyr, estimatr,
  lubridate, knitr, parallel,
  lfe,
  here, magrittr, kableExtra, snakecase, janitor, lubridate,
  data.table, knitr, jtools, huxtable, estimatr, haven, ipumsr
)
# Define pink color
red_pink <- "#e64173"
turquoise <- "#20B2AA"
orange <- "#FFA500"
red <- "#fb6107"
blue <- "#2b59c3"
green <- "#8bb174"
grey_light <- "grey70"
grey_mid <- "grey50"
grey_dark <- "grey20"
purple <- "#6A5ACD"
slate <- "#314f4f"
# Dark slate grey: #314f4f
# Knitr options
opts_chunk$set(
  comment = "#>",
  fig.align = "center",
  fig.height = 7,
  fig.width = 10.5,
  warning = F,
  message = F
)
opts_chunk$set(dev = "svg")
options(device = function(file, width, height) {
  svg(tempfile(), width = width, height = height)
})
options(crayon.enabled = F)
options(knitr.table.format = "html")
# A blank theme for ggplot
theme_empty <- theme_bw() + theme(
  line = element_blank(),
  rect = element_blank(),
  strip.text = element_blank(),
  axis.text = element_blank(),
  plot.title = element_blank(),
  axis.title = element_blank(),
  plot.margin = structure(c(0, 0, -0.5, -1), unit = "lines", valid.unit = 3L, class = "unit"),
  legend.position = "none"
)
theme_simple <- theme_bw() + theme(
  line = element_blank(),
  panel.grid = element_blank(),
  rect = element_blank(),
  strip.text = element_blank(),
  axis.text.x = element_text(size = 18, family = "STIXGeneral"),
  axis.text.y = element_blank(),
  axis.ticks = element_blank(),
  plot.title = element_blank(),
  axis.title = element_blank(),
  # plot.margin = structure(c(0, 0, -1, -1), unit = "lines", valid.unit = 3L, class = "unit"),
  legend.position = "none"
)
theme_axes_math <- theme_void() + theme(
  text = element_text(family = "MathJax_Math"),
  axis.title = element_text(size = 22),
  axis.title.x = element_text(hjust = .95, margin = margin(0.15, 0, 0, 0, unit = "lines")),
  axis.title.y = element_text(vjust = .95, margin = margin(0, 0.15, 0, 0, unit = "lines")),
  axis.line = element_line(
    color = "grey70",
    size = 0.25,
    arrow = arrow(angle = 30, length = unit(0.15, "inches")
  )),
  plot.margin = structure(c(1, 0, 1, 0), unit = "lines", valid.unit = 3L, class = "unit"),
  legend.position = "none"
)
theme_axes_serif <- theme_void() + theme(
  text = element_text(family = "MathJax_Main"),
  axis.title = element_text(size = 22),
  axis.title.x = element_text(hjust = .95, margin = margin(0.15, 0, 0, 0, unit = "lines")),
  axis.title.y = element_text(vjust = .95, margin = margin(0, 0.15, 0, 0, unit = "lines")),
  axis.line = element_line(
    color = "grey70",
    size = 0.25,
    arrow = arrow(angle = 30, length = unit(0.15, "inches")
  )),
  plot.margin = structure(c(1, 0, 1, 0), unit = "lines", valid.unit = 3L, class = "unit"),
  legend.position = "none"
)
theme_axes <- theme_void() + theme(
  text = element_text(family = "Fira Sans Book"),
  axis.title = element_text(size = 18),
  axis.title.x = element_text(hjust = .95, margin = margin(0.15, 0, 0, 0, unit = "lines")),
  axis.title.y = element_text(vjust = .95, margin = margin(0, 0.15, 0, 0, unit = "lines")),
  axis.line = element_line(
    color = grey_light,
    size = 0.25,
    arrow = arrow(angle = 30, length = unit(0.15, "inches")
  )),
  plot.margin = structure(c(1, 0, 1, 0), unit = "lines", valid.unit = 3L, class = "unit"),
  legend.position = "none"
)
theme_set(theme_gray(base_size = 20))
# Column names for regression results
reg_columns <- c("Term", "Est.", "S.E.", "t stat.", "p-Value")
# Function for formatting p values
format_pvi <- function(pv) {
  return(ifelse(
    pv < 0.0001,
    "<0.0001",
    round(pv, 4) %>% format(scientific = F)
  ))
}
format_pv <- function(pvs) lapply(X = pvs, FUN = format_pvi) %>% unlist()
# Tidy regression results table
tidy_table <- function(x, terms, highlight_row = 1, highlight_color = "black", highlight_bold = T, digits = c(NA, 3, 3, 2, 5), title = NULL) {
  x %>%
    tidy() %>%
    select(1:5) %>%
    mutate(
      term = terms,
      p.value = p.value %>% format_pv()
    ) %>%
    kable(
      col.names = reg_columns,
      escape = F,
      digits = digits,
      caption = title
    ) %>%
    kable_styling(font_size = 20) %>%
    row_spec(1:nrow(tidy(x)), background = "white") %>%
    row_spec(highlight_row, bold = highlight_bold, color = highlight_color)
}
```

```{css, echo = F, eval = F}
@media print {
  .has-continuation {
    display: block !important;
  }
}
```


## Today

- `for()` Loops
- Functional Programming
- Iterate over columns in a data frame
- Handling the output
- Conditionals and functions
- `dplyr` vectorized conditionals

---

layout: true

# Introduction to Iteration
---

<br>

Suppose you're writing code and then you realize you need to take the same code chunk and copy-paste it a bunch of times to do the same process with different inputs. 

--

Maybe you just have a handful of those inputs, in which case copy-paste isn't elegant, but it works. Or maybe you have a ton of inputs, where copy-paste would take all day!

--

The solution is to **iterate** your process over your inputs. First we'll learn iteration with `for` loops, then with `purrr::map`, and then we'll practice.

---

layout: true

# for() Loops

---

```{r for_format, eval = F}
# For each variable in a sequence, execute the expression:
for(variable in sequence){
  expr
}
```

For example, we can take a character vector and use a `for` loop and `nchar()` to print the number of characters in each string.
```{r for_cities, eval = F}
cities <- c("New York", "Paris", "London")

# The copy-paste version of this would be:
nchar(cities[1])
nchar(cities[2])
nchar(cities[3])

# The for loop version is:
for(city in cities){
  print(nchar(city))
}
```


---

```{r for_cities1}
cities <- c("New York", "Paris", "London")

# The for loop version is:
for(city in cities){
  print(nchar(city))
}
```

`city` is our variable, and during each iteration we print the number of characters of the `city`. The for loop stops when it's iterated through the whole sequence (there are no more elements of `cities`).

---

#### Example: loop over a vector

```{r for_exercise1}
# Here we write a for() loop that iterates over a vector and adds 5.
  # Here's the vector:
my_vector <- c(8, 1, 4, NA, 18)

# The copy-paste version is:
#my_vector[1] + 5
#my_vector[2] + 5
#my_vector[3] + 5
#my_vector[4] + 5
#my_vector[5] + 5

# Use x to fill in the `for` loop.
  # Hint: you can call the `var` anything you want, just be consistent.
for(x in my_vector){print(x + 5)}
```

---

The exercises so far are pretty silly because `nchar` and `+` are already **vectorized**. Function can take a vector as input and knows how to iterate:

```{r for_vectorized}
cities <- c("New York", "Paris", "London")
my_vector <- c(8, 1, 4, NA, 18)

nchar(cities)
my_vector + 5
```

More useful `for` loop: `read.csv` is a function that imports data files in "comma separated value" format. It's not vectorized: like `library`, one input at a time. Can write a `for` loop to repeat `read.csv` over multiple inputs!

---

<br>

#### Loading data with a loop

```{r load_Data, eval=FALSE}
vector_csv <- c("dataset_1.csv", "dataset_2.csv", "dataset_3.csv",
                "dataset_4.csv", "dataset_5.csv", "dataset_6.csv")

# The copy-paste version:
read.csv("dataset_1.csv")
read.csv("dataset_2.csv")
read.csv("dataset_3.csv")
read.csv("dataset_4.csv")
read.csv("dataset_5.csv")
read.csv("dataset_6.csv")

# The for loop version:
for(x in vector_csv){read.csv(x)}

```

---

<br>

#### Construct a series

```{r}
# Create a draw of 1,000 iid error terms, N(0,1)
obs    = 1000  # Sample size
vareps = rnorm(obs, mean=0, sd=1) 
X      = sample(1:8, obs, replace=TRUE)
beta   = 2

# For loop
y = c()  # Create an empty vector to fill

for(t in 1:obs){
  y[t] = beta * X[t] + vareps[t] 
}
```

---

layout: true

# Functional Programming

---

<br>

Loop constructs can be slow. Avoiding with FP is a common option. FP iterates over a vector or list of inputs and "maps" to a list of outputs using a function. Two ways to implement FP in R:

--

1. The `base` package included with R provides the `*apply()` family of functions.

1. Package `purrr` (in tidyverse) provides `map*()` family of functions. Name comes from "making your functions purrr". 

--

Here is the [Purrr CheatSheet](https://github.com/rstudio/cheatsheets/blob/master/purrr.pdf) for a snapshot of how to apply its functions.

---


`lapply`: Create a list of results, using a sequence through a specific function


```{r }
L= lapply(1:3, function(i) {
  df = tibble(num = i, let = LETTERS[i])
  return(df)
}) %>% bind_rows()
```

--

`sapply`: Create an object with results in the form of the input<br> (1:3 is vector $\implies$ $2\times 3$ matrix)

```{r}
S =sapply(1:3, function(i) {
  df = tibble(num = i, let = LETTERS[i])
  return(df)
})
```

---

```{r map_cities}
p_load(tidyverse)
cities <- c("New York", "Paris", "London")

# Recall, we want the number of characters in each element of this vector.
#nchar(cities[1])
#nchar(cities[2])
#nchar(cities[3])

# So our function is `nchar` and we want to iterate over `cities`.
  # `map()` is very elegant. It takes arguments like this: map(.x, .f), where
    # .x is the object you want to iterate over, and
    # .f is the function you want to apply.
map(cities, nchar)
```

---

Shifting to `map*()` family, pretty much the same syntax as `lapply()`.

--

These variants are useful for returning objects of a desired type. For example, we can use `purrr::map_df()` to return a data frame.

```{r map_isna, eval = F}
# map() over my_vector with the function `is.na`.
  # remember: map(.x, .f)
my_vector <- c(8, 1, 4, NA, 18)
map(my_vector, is.na)
```

---

```{r}
map_df(1:6, function(i){ df = tibble(A= i+5, B = i^2 - 3)})
```


---

Using `map()`, it'll often be useful to refer to the input `.x` when you're defining the function `.f`. For example, say we want to use `map()` to add 5 to every element of my_vector:

```{r map_formula, eval = F}
# Refer to the object you're iterating over by making the function a "formula"
  # Start the function with a `~` (that indicates a formula is coming up)
  # Refer to the iteration object with `.x`: that's the argument name
map(my_vector, ~ .x + 5)
```

Now you try:

---

#### Example: Map formulas
```{r map_exercise1, eval = F}
my_vector <- c(8, 1, 4, NA)

#Iterate over every element of my_vector to check if the element is an even number.
  # Hint: use the modulo operator %% to get the remainder after repeated division
    # If a number is even, (x %% 2) == 0 will be TRUE
    # If a number is odd, (x %% 2) == 0 will be FALSE
map()
```

---

#### Example: Map formulas
```{r map_exercise1_1}
my_vector <- c(8, 1, 4, NA)

#Iterate over every element of my_vector to check if the element is an even number.
  # Hint: use the modulo operator %% to get the remainder after repeated division
    # If a number is even, (x %% 2) == 0 will be TRUE
    # If a number is odd, (x %% 2) == 0 will be FALSE
map(my_vector, ~ (.x %% 2) == 0)
```


---

layout: true

# Iterate over columns in a data frame

---

Now you know how to use `for` loops and `purrr::map` to iterate a process over a vector of inputs.

In practice though, you'll often need to iterate a process over many **columns** of inputs in a data frame. Think about in `gapminder`, if we wanted to take the `mean` of the `life_exp` column, the `gdp_per_cap` column, and then the `pop` column. If we had a data frame with 100 columns, we would want to find a way to iterate that process instead of copy-pasting `mean(column1)`, `mean(column2)` over and over. 

We'll do an example (using both a `for` loop and `map`) where we iterate over the `gapminder` columns and draw scatterplots that visualize those variables against `gapminder$year`.

---

#### `for` loop in a data frame

```{r for_columns, eval = F}
# Let's remind ourselves what `gapminder` looks like:
head(gapminder)

# Here's the copy-paste version of what we'd like to do:
ggplot(gapminder) +
  geom_point(aes(x = year, y = life_exp))
ggplot(gapminder) +
  geom_point(aes(x = year, y = gdp_per_cap))
ggplot(gapminder) +
  geom_point(aes(x = year, y = pop))
```

---

```{r Checkdata}
p_load(gapminder)
head(gapminder)
```

---

Doing the `for` loop instead of the copy-pasta solution:

```{r for_columns2111, eval=FALSE}
# What are we iterating across? Columns in `gapminder`. 
#I'll call the variable `cols`, but that's completely arbitrary.
p_load(gapminder)
for(cols in gapminder){
  print(
    ggplot(gapminder) +
      geom_point(aes(x = year, y = cols))  )  }
```

Notice it will even plotted `country` and `continent` against year. But how did it know to iterate over the columns of gapminder rather than the individual elements or the rows?

R knew that we wanted to iterate over columns because that's the default behavior when you iterate over a 2-dimensional data frame. This is a big part of the reason we try to keep data in this vertical format with variables as column names and observations as rows.

---

#### Example: `for` loop over a data frame
```{r for_exercise2, eval = F}
simple_data_frame <- tibble(
  column1 = c(1, 2, 3, 4),
  column2 = c(5, 6, 7, 8),
  column3 = c(9, 10, 11, 12)
)

# Loop over the data frame and print the `mean` of each column

for(x in simple_data_frame){
  
}
```

---

#### Example: `for` loop over a data frame
```{r for_exercise21}
simple_data_frame <- tibble(
  column1 = c(1, 2, 3, 4),
  column2 = c(5, 6, 7, 8),
  column3 = c(9, 10, 11, 12)
)

# Loop over the data frame and print the `mean` of each column

for(x in simple_data_frame){
  print(mean(x))
}
```

---

Now we'll go back to looping over `gapminder` columns, except now we'll use `purrr::map` to solve the problem.

```{r map_columns, eval = F}
# map(.x, .f)
  # .x: What are we iterating across? `gapminder` columns. We can actually take 
  # advantage of R's default behavior with 2-D data frames and just iterate over 
  #`gapminder`, R will know what to do. So `.x = gapminder`.
  # .f: What's the function? Drawing this ggplot. It would be complicated to stuff 
  # all that code in the map() call, so we'll write our own function first.
    # Note, this is the format for writing your own function:

func_name <- function(arg1, arg2){
  "func_body_here"
}

draw_yearly_scatterplot <- function(variable){
    ggplot(gapminder) +
      geom_point(aes(x = year, y = variable))
}

map(gapminder, draw_yearly_scatterplot)
```


---

#### Example: `map()` over a data frame
```{r map_simpledf}
simple_data_frame <- tibble(
  column1 = c(1, 2, 3, 4),
  column2 = c(5, 6, 7, 8),
  column3 = c(9, 10, 11, 12)
)

# map() over the data frame and find the `mean` of each column
  # Recall, R's default behavior is to map over columns

map(simple_data_frame, mean)
```

---

layout: true

# Handling the Output

---

The `for` loops we've done create different output types versus when we `map`.

```{r iteration_output, eval = F}
simple_data_frame <- tibble(
  column1 = c(1, 2, 3, 4),
  column2 = c(5, 6, 7, 8),
  column3 = c(9, 10, 11, 12)   )

# print() just prints to the console. It doesn't save the output anywhere.
for(x in simple_data_frame){
  print(mean(x))   }

# map() on the other hand returns a list. You can save that output.
map(simple_data_frame, mean)
means_list <- map(simple_data_frame, mean)
# Note, if you don't want your output as a list, map() has siblings:
  # map_dbl() returns a double (numeric) vector, 
  # map_df() returns a data frame
  # there are many others
```


---

How do we save the results of a `for` loop? The solution is to initialize an empty vector before the loop, and inside the `for` loop, during each iteration, `append()` the results onto the vector instead of printing.

```{r for_save}
means_vec <- vector()

for(x in simple_data_frame){
  means_vec <- append(means_vec, mean(x))
}

means_vec
```

---

So now you know how to:

* `for` loop over a vector of inputs
* `purrr::map` over a vector of inputs
* `for` loop over columns of a data frame
* `purrr::map` over columns of a data frame
* And save those results!

Next up: More function techniques using `if` statements and `if else` ladders.

***Resources:***

[Datacamp Foundations of Functional Programming with purrr](https://learn.datacamp.com/courses/foundations-of-functional-programming-with-purrr)<br>
[Datacamp Introduction to Writing Functions in R](https://learn.datacamp.com/courses/introduction-to-writing-functions-in-r)

---

layout: true

# Conditionals and Functions

---

<br>

In these latter slides we'll cover **conditionals**:

  * `if` statements and `if else` ladders
  * `dplyr` vectorized conditionals `if_else` and `case_when`
  
Then we'll talk more about **writing functions**.

---

<br>

`if` and `if else`: Conditional statements do actions *conditioned* on a logical statement being TRUE.

This is the format for an `if` statement:
```{r if_format, eval = F}
if(condition){
  expr  }
```

For example,
```{r if_1expr, eval = F}
x <- -3

if(x < 0){
  print("x is a negative number")   }
```

---

<br>

Perhaps we want to check only a specific set of values to another value. Let's do a very simple case of this first. It will change a to 5 if x=3 is true.

```{r if_ex1check}
x <- 3
a <- 2

if(x == 3){
  a <- 5
}

a
```


---

We can add an `else`:
```{r ifelse_format, eval = F}
if(condition){
  expr1
} else {
  expr2
}
```

For example,
```{r ifelse_2expr, eval = F}
x <- 5

if(x < 0){
  print("x is a negative number")
} else {
  print("x is not a negative number")
}
```

---

<br>

We can nest if else statements to deal with 3 or more expressions. Parentheses are not needed:

.pull-left[
```{r ifelse_ladder_format, eval = F}
if(condition1){expr1} 
else if(condition2){expr2}
else{expr3}
```
]

.pull-right[
```{r ifelse_3expr, eval = F}
x <- 0
  
if(x < 0){
  print("x is a negative number")
} else if(x == 0) {
  print("x is zero")
} else{
  print("x is a positive number")
}
```
]


---

```{r for_if}
Y_0 = c(2, 2.5, 4, 1.5, 3) #untreated outcomes
Y_1 = c(5, 3, 2, 4, 1) #treated outcomes
D = c(1, 0, 1, 0, 1) #treatment status

# Could use `purrr::map` but let's write a `for` loop with if inside.
# Keep in mind, there's a way easier way to do this that we'll do next.

Y <- vector() #initialize an empty vector

for(i in 1:length(D)){
  d <- D[i]
 if(d == 1){
   Y <- append(Y, Y_1[i])
 } else {
   Y <- append(Y, Y_0[i])
 }
}

Y #check
```

---

layout: true

# dplyr vectorized conditionals

---

Recall our first iteration was;
```{r for_cities2, eval = F}
cities <- c("New York", "Paris", "London")

# The for loop version is:
for(city in cities){
  print(nchar(city))}
```

Was silly because `nchar()` is vectorized, which means if you input a vector, the function already understands it should iterate across that vector.

```{r nchar_vectorized, eval = F}
nchar(cities)
```

So if you need to iterate, you can:

* use a `for` loop
* use `purrr::map`
* or if a vectorized function exists, just use that! 

---

<br>

Lots of functions in base R and in the tidyverse are vectorized.
`dplyr` has two functions `if_else` and `case_when` that are vectorized, but they're so easy, I like to use them whether or not I need to iterate over a vector. They can do everything from Section 5 on single values or on vectors.

The format for `if_else` is:
```{r if_else_format, eval = F}
if_else(condition, true, false)

# true: value to use if the condition is TRUE
# false: value to use if the condition is FALSE
```

---

<br>

For example,
```{r if_else_2expr}
x <- c(5, 0, -5)
if_else(x >= 0, "x is non-negative", "x is negative")
```

Suppose we wanted to create an indicator for whether or not an observation is in Europe and has a gdp_per_cap greater than 20,000:
```{r if_else_gapminder, eval = F}
p_load(gapminder)
if_else(gapminder$continent == "Europe" & gapminder$gdpPercap > 20000, 1, 0)
#Not printing since its a crazy output of 0's and 1's.
```

---

<br>

#### Example: `if_else`
```{r if_else_ex1, eval = F}
Y_0 = c(2, 2.5, 4, 1.5, 3) #untreated outcome
Y_1 = c(5, 3, 2, 4, 1) #treated outcome
D = c(1, 0, 1, 0, 1) #treatment status

# Write an `if_else` statement that creates variable Y: 
# where if D == 1, let Y be Y_1; if D == 0, let Y be Y_0.
```

---

<br>

#### Example: `if_else`
```{r if_else_ex_1, eval = F}
Y_0 = c(2, 2.5, 4, 1.5, 3) #untreated outcome
Y_1 = c(5, 3, 2, 4, 1) #treated outcome
D = c(1, 0, 1, 0, 1) #treatment status

# Write an `if_else` statement that creates variable Y: 
# where if D == 1, let Y be Y_1; if D == 0, let Y be Y_0.

Y <- if_else(D == 1, Y_1, Y_0)

# This is one line of code! Previously, we had to use 10 lines 
# to do this same thing using `for` and `if`.
```

---

If you have 3 expressions instead of just 2, you *could* nest an `if_else` inside an `if_else`, but it's pretty confusing to read:

```{r case_when, eval = F}
x <- c(5, 0, -5)
if_else(x > 0, "x is positive", if_else(x == 0, "x is zero", "x is negative"))
```

Instead, use `case_when`:

```{r case_when_3expr, eval = F}
x <- c(5, 0, -5)
case_when(
  x > 0 ~ "x is positive",
  x == 0 ~ "x is zero",
  x < 0 ~ "x is negative"
)
```


---

exclude: true

```{R generate pdfs, include = F, eval = F}
#remotes::install_github('rstudio/pagedown')
library(pagedown)
pagedown::chrome_print("Tutorial-4.html", output = "Tutorial-4.pdf")
#pagedown::chrome_print("Tutorial Slides 1-nopause.html", output = "01-research-r-nopause.pdf")
```