---
title: "Visualization: Using ggplot2"
subtitle: "EC 607 Metrics, Tutorial 3"
author: "Philip Economides"
date: "Spring 2021"
output:
  xaringan::moon_reader:
    css: ['default', 'metropolis', 'metropolis-fonts', 'my-css.css']
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---
class: middle

```{r CSS, eval=FALSE, include=FALSE}
#css: ['default', 'metropolis', 'metropolis-fonts', 'my-css.css']
```

```{R, setup, include = F}
# devtools::install_github("dill/emoGG")
library(pacman)
p_load(
  broom, tidyverse,
  latex2exp, ggplot2, ggthemes, ggforce, viridis, extrafont, gridExtra,
  kableExtra, snakecase, janitor,
  data.table, dplyr, estimatr,
  lubridate, knitr, parallel,
  lfe,
  here, magrittr
)
# Define pink color
red_pink <- "#e64173"
turquoise <- "#20B2AA"
orange <- "#FFA500"
red <- "#fb6107"
blue <- "#2b59c3"
green <- "#8bb174"
grey_light <- "grey70"
grey_mid <- "grey50"
grey_dark <- "grey20"
purple <- "#6A5ACD"
slate <- "#314f4f"
# Dark slate grey: #314f4f
# Knitr options
opts_chunk$set(
  comment = "#>",
  fig.align = "center",
  fig.height = 7,
  fig.width = 10.5,
  warning = F,
  message = F
)
opts_chunk$set(dev = "svg")
options(device = function(file, width, height) {
  svg(tempfile(), width = width, height = height)
})
options(crayon.enabled = F)
options(knitr.table.format = "html")
# A blank theme for ggplot
theme_empty <- theme_bw() + theme(
  line = element_blank(),
  rect = element_blank(),
  strip.text = element_blank(),
  axis.text = element_blank(),
  plot.title = element_blank(),
  axis.title = element_blank(),
  plot.margin = structure(c(0, 0, -0.5, -1), unit = "lines", valid.unit = 3L, class = "unit"),
  legend.position = "none"
)
theme_simple <- theme_bw() + theme(
  line = element_blank(),
  panel.grid = element_blank(),
  rect = element_blank(),
  strip.text = element_blank(),
  axis.text.x = element_text(size = 18, family = "STIXGeneral"),
  axis.text.y = element_blank(),
  axis.ticks = element_blank(),
  plot.title = element_blank(),
  axis.title = element_blank(),
  # plot.margin = structure(c(0, 0, -1, -1), unit = "lines", valid.unit = 3L, class = "unit"),
  legend.position = "none"
)
theme_axes_math <- theme_void() + theme(
  text = element_text(family = "MathJax_Math"),
  axis.title = element_text(size = 22),
  axis.title.x = element_text(hjust = .95, margin = margin(0.15, 0, 0, 0, unit = "lines")),
  axis.title.y = element_text(vjust = .95, margin = margin(0, 0.15, 0, 0, unit = "lines")),
  axis.line = element_line(
    color = "grey70",
    size = 0.25,
    arrow = arrow(angle = 30, length = unit(0.15, "inches")
  )),
  plot.margin = structure(c(1, 0, 1, 0), unit = "lines", valid.unit = 3L, class = "unit"),
  legend.position = "none"
)
theme_axes_serif <- theme_void() + theme(
  text = element_text(family = "MathJax_Main"),
  axis.title = element_text(size = 22),
  axis.title.x = element_text(hjust = .95, margin = margin(0.15, 0, 0, 0, unit = "lines")),
  axis.title.y = element_text(vjust = .95, margin = margin(0, 0.15, 0, 0, unit = "lines")),
  axis.line = element_line(
    color = "grey70",
    size = 0.25,
    arrow = arrow(angle = 30, length = unit(0.15, "inches")
  )),
  plot.margin = structure(c(1, 0, 1, 0), unit = "lines", valid.unit = 3L, class = "unit"),
  legend.position = "none"
)
theme_axes <- theme_void() + theme(
  text = element_text(family = "Fira Sans Book"),
  axis.title = element_text(size = 18),
  axis.title.x = element_text(hjust = .95, margin = margin(0.15, 0, 0, 0, unit = "lines")),
  axis.title.y = element_text(vjust = .95, margin = margin(0, 0.15, 0, 0, unit = "lines")),
  axis.line = element_line(
    color = grey_light,
    size = 0.25,
    arrow = arrow(angle = 30, length = unit(0.15, "inches")
  )),
  plot.margin = structure(c(1, 0, 1, 0), unit = "lines", valid.unit = 3L, class = "unit"),
  legend.position = "none"
)
theme_set(theme_gray(base_size = 20))
# Column names for regression results
reg_columns <- c("Term", "Est.", "S.E.", "t stat.", "p-Value")
# Function for formatting p values
format_pvi <- function(pv) {
  return(ifelse(
    pv < 0.0001,
    "<0.0001",
    round(pv, 4) %>% format(scientific = F)
  ))
}
format_pv <- function(pvs) lapply(X = pvs, FUN = format_pvi) %>% unlist()
# Tidy regression results table
tidy_table <- function(x, terms, highlight_row = 1, highlight_color = "black", highlight_bold = T, digits = c(NA, 3, 3, 2, 5), title = NULL) {
  x %>%
    tidy() %>%
    select(1:5) %>%
    mutate(
      term = terms,
      p.value = p.value %>% format_pv()
    ) %>%
    kable(
      col.names = reg_columns,
      escape = F,
      digits = digits,
      caption = title
    ) %>%
    kable_styling(font_size = 20) %>%
    row_spec(1:nrow(tidy(x)), background = "white") %>%
    row_spec(highlight_row, bold = highlight_bold, color = highlight_color)
}
```

```{css, echo = F, eval = F}
@media print {
  .has-continuation {
    display: block !important;
  }
}
```

name: schedule

# Schedule

## Today

- An introduction to visualization.
- How to build a plot.
- Plot components.
- Experiment with `ggplot2`.
- `ggplot2` resources.

---
layout: true

# Introduction
---
name: basics

## What is it?

`ggplot2` is a package included in the `tidyverse`. It lets you easily make plots that are extremely customizable through the layering of graphics. 

```{r ggplot_layers, eval = F}
#Load your Packages
if (!require("pacman")) install.packages("pacman")
pacman::p_load(gapminder, tidyverse)

summary(gapminder)

# Layered graphics: scatterplot + linear model
ggplot(gapminder, mapping = aes(x = log(gdpPercap), y = lifeExp)) + 
  geom_point() +
  geom_smooth(method = lm)
```

--
This section is based on a paper [A Layered Grammar of Graphics](https://vita.had.co.nz/papers/layered-grammar.pdf) by Hadley Wickham, one of the creators of ggplot2.

---
layout: true

# Basic Plot

---
## How to build a plot

First we need some data.

`ggplot2` requires that data be in the form of a data frame. If your data is not, sometimes you can coerce: `as_data_frame()` actually makes your data a tibble (equivalently, as_tibble()), which is the tidyverse's updated version of a data frame.

```{r ggplot_simple_data}
# A tribble is just a transposed tibble: for convenience, I've decided 
#to define it row-wise instead of column-wise.

simple_data <- tribble(
  ~"A", ~"B", ~"C", ~"D",
  #---|-----|-----|-----
  2, 3, 4, "b",
  1, 2, 1, "b",
  4, 5, 15, "a",
  9, 10, 80, "a")
```
--
This dataset has 4 variables (A, B, C, D), and 4 observations of those variables (4 rows). 

---

## A basic plot

Let's make a scatterplot of A versus C, and draw different shapes for realizations of D. When we're building this plot, we're hoping that behind the scenes, R creates a new dataset with these changes:

* Map the x-position to A
* Map the y-position to C
* Map shape to D
* Remove any variables that don't appear in the plot

These mappings (x-position, y-position, and shape) are examples of **aesthetics**, `aes` for short.

---

Behind the scenes, `ggplot2` will take our `simple_data` and create a new dataset to help build our plot:

```{r ggplot_simple_data_map, eval = F}
simple_data_mapping <- tribble(
    ~"x", ~"y", ~"Shape",
    #---|-----|---------
    2, 4, "triangle",
    1, 1, "triangle",
    4, 15, "circle",
    9, 80, "circle"
)

# Compare `simple_data` to `simple_data_mapping`:
simple_data
simple_data_mapping

ggplot(simple_data) + 
  geom_point(mapping = aes(x = A, y = C, shape = D))

# Note: inside a `ggplot`, we use `+` instead of `%>%`.
```

---

```{r CrashParty, fig.width = 6, fig.height = 4}

ggplot(simple_data) + 
  geom_point(mapping = aes(x = A, y = C, shape = D))

```

---

layout: true

#Faceting

---



- Notice, `ggplot` mapped "a" to a circle and "b" to a triangle. This transformation is the responsibility of **scales**. So if you want the mapping to be different, specify a different scale.

- We can create lots of different types of plots using this same basic specification. For example, if we drew lines instead of points, we would get a line plot. 

- Or, if we used bars, we would get a bar plot. Bars, lines, and points are all examples of **geometric objects**, `geom` for short. 

- `ggplot2` has a [ton of options for geoms](https://ggplot2.tidyverse.org/reference/), but we'll be using `geom_point`, `geom_line`, `geom_bar`, etc.

- Let's try more complicated plot: Faceting.

---

Faceting your plot into groups specified by D. 

Each value of D (a or b) will be displayed on a different panel.

```{r ggplot_facet, fig.width = 6, fig.height = 3.5}
ggplot(simple_data) + 
  geom_point(mapping = aes(x = A, y = C, shape = D)) + 
  facet_wrap(~D)

# Notice that shape = D is redundant: we're displaying different values for D in 
# different panels, so it's a little silly to also emphasize different values for D with shape.
```

---

For example, in our `gapminder`,`log(gdpPercap)` has a high correlation with `lifeExp` across continents.

```{r ggplot_facet_ex, fig.width=6, fig.height=3.5}
library(gapminder)
ggplot(gapminder) + 
  geom_point(mapping = aes(x = log(gdpPercap), y = lifeExp)) +
  facet_wrap(~continent)
```

---

layout: true

#Plot components

---

Components that make up a plot:

* data and aesthetic mappings
* geometric objects
* scales
* facet specification
* statistical transformations
* coordinate system

Together, the data, mappings, statistical transformation, and geometric object form a **layer**. A plot can have many layers. For example, we can overlay a scatterplot with a fitted line. Or you can draw two curves on top of each other from different data sets.

---

A **statistical transformation** (`stat` for short) typically summarizes the data in some manner. One useful stat is the smoother: it calculates the mean of y conditional on x, subject to some restriction that ensures smoothness.

```{r ggplot_stat_smooth, fig.height=3, fig.width=4.5}
ggplot(gapminder, mapping = aes(x = log(gdpPercap), y = lifeExp)) + 
  geom_point(alpha=0.5) +
  geom_smooth()
```

---

Another example of statistical transformation: aggregation of `geom_histogram`. Only specify x variable. R creates the y variable by dividing the range of x by the number of `bins` you specify (30 by default), and then counting number of observations in each bin.

```{r ggplot_stat_hist, fig.height=3.3}
# Default bins is 30
ggplot(gapminder, mapping = aes(x = gdpPercap)) + 
  geom_histogram()
```

---

By specifying the number of `bins` we can introduce more detail into the histogram.

--

```{r ggplot_stat_histband, fig.height=3.5}
# Specify binwidth to be 1,000 (or `bins = 50`, or `breaks` = ...)
ggplot(gapminder, mapping = aes(x = gdpPercap)) + 
  geom_histogram(binwidth = 1000)
```

--

To add dollar signs and commas, we'll need `scales` package. 

---

layout: true

#Scale

---

Every aesthetic mapping has a scale.
--

* Shape: circle, square, triangle, etc
* Size: how big or small should the points be?
* Color: the default color scheme in ggplot2 picks evenly spaced hues around the hcl (Hue-Chroma-Luminance) color wheel starting from 15. So if you have 3 categories for variables, you'll get different colors compared to if you have 5. If your variable is continuous, your scale is a path in the HCL color space.

--

Here are some legends associated with [scales](https://encrypted-tbn0.gstatic.com/images?q=tbn%3AANd9GcRy3CB_vS1GJ51iMmKBY6JCMD0Ev1GpcdIAvnRtSxRiaK_k-i-K&usqp=CAU)

I'll also add some links in order to access a [color palette generator](https://learnui.design/tools/data-color-picker.html) and [color encyclopedia](https://www.colorhexa.com/). 

---

```{r ggplot_fixed_val, fig.height=5}
# If you want to set an aesthetic of your plot to be some fixed value,
# make that specification outside the aesthetic mapping:

ggplot(gapminder, aes(x = log(gdpPercap), y = lifeExp)) +
  geom_point(color = "blue")
```


---
Lets colour by groups!

```{r ggplot_fixed_val_ex, fig.height=5}
# Make point color = continent, size = 2, and shape = "k"
ggplot(gapminder, aes(x = log(gdpPercap), y = lifeExp)) + 
  geom_point(aes(color = continent), size = 2, shape = "k")
```


---

layout:true

# Exploring ggplot2

---

Consider the following base code. 

```{r base-plot, eval = F}
gapminder %>%
  ggplot(mapping = aes(x = log(gdp_per_cap), y = life_exp)) +
  geom_point(aes(color = continent))
```

How can we...

- Add a title and re-name the legend
- Change the color scale from the default
- Adjust transparency of points
- Represent population by point `size`
- Change the geom from `geom_point` to `geom_text`
- Make a line plot showing how lifeExp has changed over time for a few countries
- Make a bar plot or histogram to see data frequency

---

**1. Add a title and re-name the legend:**

Add: `labs(title = "___", color = "___")`.

```{r plot-titles, fig.height=4}
gapminder %>%
  ggplot(mapping = aes(x = log(gdpPercap), y = lifeExp)) +
  geom_point(aes(color = continent)) +
  labs(title = "GDP per capita predicts life expectancy", color = element_blank())
```

---

**2. Change the color scale from the default. Options:**

Add `scale_color_viridis_d()` (d for discrete, c for continuous variables). Contrasts are preserved when printing in black and white, colorblind-safe.
  
```{r plot-viridis, fig.height=3.8}
gapminder %>%
  ggplot(mapping = aes(x = log(gdpPercap), y = lifeExp)) +
  geom_point(aes(color = continent)) + scale_color_viridis_d()
```

---

Use Wes Anderson movie color palettes with the package `wesanderson`. Check them out for yourself on [github](https://github.com/karthik/wesanderson). 

```{r plot-wesanderson, fig.height=3.8}
p_load(wesanderson)
gapminder %>%
  ggplot(mapping = aes(x = log(gdpPercap), y = lifeExp)) +
  geom_point(aes(color = continent)) +
  scale_color_manual(values=wesanderson::wes_palette(n=5, name= "FantasticFox1"))
```

---

Pick your own, `scale_color_manual(values = c("__", "__", ...))`. Pick 5 colors for 5 continents in the data.
  
```{r plot-manual-colors, fig.height=3.8}
gapminder %>%
  ggplot(mapping = aes(x = log(gdpPercap), y = lifeExp)) +
  geom_point(aes(color = continent)) +
  scale_color_manual(values = c("chartreuse3", "dodgerblue2", 
                                "maroon3", "gold2", "wheat1"))
```

---

**3. Adjust transparency of points** 

Do so by setting `alpha` to a fixed value between 0 and 1 (outside the aesthetic mapping). This helps when there's overplotting like below.

```{r plot-alpha, fig.height=3.8}
gapminder %>%
  ggplot(mapping = aes(x = log(gdpPercap), y = lifeExp)) +
  geom_point(aes(color = continent), alpha = .3)
```

---

**4. Represent population by point `size`** 

Specify `size = pop` in the aesthetic mapping 

```{r plot-point-size, fig.height=4.5}
gapminder %>%
  ggplot(mapping = aes(x = log(gdpPercap), y = lifeExp)) +
  geom_point(aes(color = continent, size = pop))
```

---

```{r plot-point-format, fig.height=4.5}
gapminder %>%
  ggplot(mapping = aes(x = log(gdpPercap), y = lifeExp)) +
  geom_point(aes(color = continent, size = pop), alpha=0.5)+
  scale_size_continuous(labels = scales::unit_format(unit = "M", scale = 1e-6))+
  xlab("Log, GDP per Capita")+ ylab("Life Expectancy")+
  labs(size="Population", colour="Continent")
```

---

**5. Change the geom from `geom_point` to `geom_text`.** 

This creates a scatterplot with the name of the country instead of points! Use `size = pop` inside the aesthetic mapping so you're sure to be able to read large country names.

```{r plot-geom-text, eval=FALSE, fig.height=4.3, include=TRUE}
gapminder %>% filter(year == 2007) %>%
  ggplot(mapping = aes(x = log(gdpPercap), y = lifeExp)) +
  geom_text(mapping = aes(label = country, size = pop, color = continent))
```

---

```{r plot-geom-txt, echo=FALSE}
gapminder %>% filter(year == 2007) %>%
  ggplot(mapping = aes(x = log(gdpPercap), y = lifeExp)) +
  geom_text(mapping = aes(label = country, size = pop, color = continent))
```


---

**6. Make a line plot showing how life_exp has changed over time for a few countries**

Use `geom_line`. Let x = year and y = life expectancy.

Re-order to match plot, `color = fct_reorder2(country, year, lifeExp, .fun = last2)`. 

`.fun = last2` to specify order by last 2 values in our data.

--

```{r plot-line, fig.height=4.8, eval=F}
p_load(forcats)
gapminder %>%
  filter(country %in% c("Singapore", "Spain", "Ireland", 
                        "Italy", "Israel", "Korea, Rep.")) %>%
  ggplot(mapping = aes(x = year, y = lifeExp, 
           color = fct_reorder2(country, year, lifeExp, .fun = last2))) +
  geom_line() + 
  scale_colour_discrete("Countries")
```

---

```{r plot-line_show, echo=F}
p_load(forcats)
gapminder %>%
  filter(country %in% c("Singapore", "Spain", "Ireland", "Italy", "Israel", "Korea, Rep.")) %>%
  ggplot(mapping = aes(x = year, y = lifeExp, color = fct_reorder2(country, year, lifeExp, .fun = last2))) +
  geom_line()+
  scale_colour_discrete("Countries")
```

---

**7. Make a bar plot or histogram to see data frequency**

Bar plots: categorical data, histograms: continuous data.<br> 
Reorder bars according to frequency: `x = fct_infreq(continent)`.
  
```{r plot-bar, fig.height=3.8}
gapminder %>% ggplot(mapping = aes(x = fct_infreq(continent))) +
  geom_bar(aes(fill = continent)) + xlab(NULL) + 
  theme(legend.position = "none")
```

 
---
  
**Histogram:** Compare the number of countries in bins of income in 2007. Add an aesthetic mapping that colors histogram by continent. Use `fill = continent` instead of `color = continent`.

```{r plot-histogram, fig.height=3.8}
gapminder %>%
  filter(year == 2007) %>%
  ggplot(mapping = aes(x = gdpPercap)) +
  geom_histogram(aes(fill = continent), bins = 10)
```

---

`ggplot2` resources 

* [A Layered Grammar of Graphics](https://vita.had.co.nz/papers/layered-grammar.pdf) 

* [R4DS](https://r4ds.had.co.nz/data-visualisation.html)

* [Datacamp](https://learn.datacamp.com/courses/data-visualization-with-ggplot2-1)

Thank you!

---

```{R, statistical languages, eval = F, fig.height = 6, fig.width = 9, dev = "svg"}
pop_df <- data.frame(
  lang = c("SQL", "Python", "R", "SAS", "Matlab", "SPSS", "Stata"),
  n_jobs = c(107130, 66976, 48772, 25644, 11464, 3717, 1624),
  free = c(T, T, T, F, F, F, F)
)
pop_df %<>% mutate(lang = lang %>% factor(ordered = T))
# Plot it
ggplot(data = pop_df, aes(x = lang, y = n_jobs, fill = free)) +
geom_col() +
geom_hline(yintercept = 0) +
aes(x = reorder(lang, -n_jobs), fill = reorder(free, -free)) +
xlab("Statistical language") +
scale_y_continuous(label = scales::comma) +
ylab("Number of jobs") +
ggtitle(
  "Comparing statistical languages",
  subtitle = "Number of job postings on Indeed.com, 2019/01/06") +
scale_fill_manual(
  "Free?",
  labels = c("True", "False"),
  values = c(red_pink, "darkslategray")) +
theme_pander(base_size = 17) +
theme(legend.position = "bottom")
```

---

```{R statistical_languages_1, echo = F, fig.height = 6, fig.width = 9, dev = "svg"}
# The popularity data
pop_df <- data.frame(
  lang = c("SQL", "Python", "R", "SAS", "Matlab", "SPSS", "Stata"),
  n_jobs = c(107130, 66976, 48772, 25644, 11464, 3717, 1624),
  free = c(T, T, T, F, F, F, F)
)
pop_df %<>% mutate(lang = lang %>% factor(ordered = T))
# Plot it
ggplot(data = pop_df, aes(x = lang, y = n_jobs, fill = free)) +
geom_col() +
geom_hline(yintercept = 0) +
aes(x = reorder(lang, -n_jobs), fill = reorder(free, -free)) +
xlab("Statistical language") +
scale_y_continuous(label = scales::comma) +
ylab("Number of jobs") +
ggtitle(
  "Comparing statistical languages",
  subtitle = "Number of job postings on Indeed.com, 2019/01/06"
) +
scale_fill_manual(
  "Free?",
  labels = c("True", "False"),
  values = c(red_pink, "darkslategray")
) +
theme_pander(base_size = 17) +
theme(legend.position = "bottom")
```



---

exclude: true

```{R generate pdfs, include = F, eval = F}
#remotes::install_github('rstudio/pagedown')
library(pagedown)
pagedown::chrome_print("Tutorial-Slides-2.html", output = "Tutorial-Slides-2.pdf")
#pagedown::chrome_print("Tutorial Slides 1-nopause.html", output = "01-research-r-nopause.pdf")
```