---
title: "Dataframes: Using dplyr"
subtitle: "EC 607 Metrics, Tutorial 2"
author: "Philip Economides"
date: "Spring 2021"
output:
  xaringan::moon_reader:
    css: ['default', 'metropolis', 'metropolis-fonts', 'my-css.css']
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---
class: inverse, middle

```{R setup, include = F}
# devtools::install_github("dill/emoGG")
library(pacman)
p_load(
  broom, tidyverse, pracma, tstools,
  latex2exp, ggplot2, ggthemes, ggforce, viridis, extrafont, gridExtra,
  kableExtra, snakecase, janitor,
  data.table, dplyr, estimatr,
  lubridate, knitr, parallel,
  lfe,
  here, magrittr, kableExtra, snakecase, janitor, lubridate,
  data.table, knitr, jtools, huxtable, estimatr, haven, ipumsr
)
# Define pink color
red_pink <- "#e64173"
turquoise <- "#20B2AA"
orange <- "#FFA500"
red <- "#fb6107"
blue <- "#2b59c3"
green <- "#8bb174"
grey_light <- "grey70"
grey_mid <- "grey50"
grey_dark <- "grey20"
purple <- "#6A5ACD"
slate <- "#314f4f"
# Dark slate grey: #314f4f
# Knitr options
opts_chunk$set(
  comment = "#>",
  fig.align = "center",
  fig.height = 7,
  fig.width = 10.5,
  warning = F,
  message = F
)
opts_chunk$set(dev = "svg")
options(device = function(file, width, height) {
  svg(tempfile(), width = width, height = height)
})
options(crayon.enabled = F)
options(knitr.table.format = "html")
# A blank theme for ggplot
theme_empty <- theme_bw() + theme(
  line = element_blank(),
  rect = element_blank(),
  strip.text = element_blank(),
  axis.text = element_blank(),
  plot.title = element_blank(),
  axis.title = element_blank(),
  plot.margin = structure(c(0, 0, -0.5, -1), unit = "lines", valid.unit = 3L, class = "unit"),
  legend.position = "none"
)
theme_simple <- theme_bw() + theme(
  line = element_blank(),
  panel.grid = element_blank(),
  rect = element_blank(),
  strip.text = element_blank(),
  axis.text.x = element_text(size = 18, family = "STIXGeneral"),
  axis.text.y = element_blank(),
  axis.ticks = element_blank(),
  plot.title = element_blank(),
  axis.title = element_blank(),
  # plot.margin = structure(c(0, 0, -1, -1), unit = "lines", valid.unit = 3L, class = "unit"),
  legend.position = "none"
)
theme_axes_math <- theme_void() + theme(
  text = element_text(family = "MathJax_Math"),
  axis.title = element_text(size = 22),
  axis.title.x = element_text(hjust = .95, margin = margin(0.15, 0, 0, 0, unit = "lines")),
  axis.title.y = element_text(vjust = .95, margin = margin(0, 0.15, 0, 0, unit = "lines")),
  axis.line = element_line(
    color = "grey70",
    size = 0.25,
    arrow = arrow(angle = 30, length = unit(0.15, "inches")
  )),
  plot.margin = structure(c(1, 0, 1, 0), unit = "lines", valid.unit = 3L, class = "unit"),
  legend.position = "none"
)
theme_axes_serif <- theme_void() + theme(
  text = element_text(family = "MathJax_Main"),
  axis.title = element_text(size = 22),
  axis.title.x = element_text(hjust = .95, margin = margin(0.15, 0, 0, 0, unit = "lines")),
  axis.title.y = element_text(vjust = .95, margin = margin(0, 0.15, 0, 0, unit = "lines")),
  axis.line = element_line(
    color = "grey70",
    size = 0.25,
    arrow = arrow(angle = 30, length = unit(0.15, "inches")
  )),
  plot.margin = structure(c(1, 0, 1, 0), unit = "lines", valid.unit = 3L, class = "unit"),
  legend.position = "none"
)
theme_axes <- theme_void() + theme(
  text = element_text(family = "Fira Sans Book"),
  axis.title = element_text(size = 18),
  axis.title.x = element_text(hjust = .95, margin = margin(0.15, 0, 0, 0, unit = "lines")),
  axis.title.y = element_text(vjust = .95, margin = margin(0, 0.15, 0, 0, unit = "lines")),
  axis.line = element_line(
    color = grey_light,
    size = 0.25,
    arrow = arrow(angle = 30, length = unit(0.15, "inches")
  )),
  plot.margin = structure(c(1, 0, 1, 0), unit = "lines", valid.unit = 3L, class = "unit"),
  legend.position = "none"
)
theme_set(theme_gray(base_size = 20))
# Column names for regression results
reg_columns <- c("Term", "Est.", "S.E.", "t stat.", "p-Value")
# Function for formatting p values
format_pvi <- function(pv) {
  return(ifelse(
    pv < 0.0001,
    "<0.0001",
    round(pv, 4) %>% format(scientific = F)
  ))
}
format_pv <- function(pvs) lapply(X = pvs, FUN = format_pvi) %>% unlist()
# Tidy regression results table
tidy_table <- function(x, terms, highlight_row = 1, highlight_color = "black", highlight_bold = T, digits = c(NA, 3, 3, 2, 5), title = NULL) {
  x %>%
    tidy() %>%
    select(1:5) %>%
    mutate(
      term = terms,
      p.value = p.value %>% format_pv()
    ) %>%
    kable(
      col.names = reg_columns,
      escape = F,
      digits = digits,
      caption = title
    ) %>%
    kable_styling(font_size = 20) %>%
    row_spec(1:nrow(tidy(x)), background = "white") %>%
    row_spec(highlight_row, bold = highlight_bold, color = highlight_color)
}
```

```{css, echo = F, eval = F}
@media print {
  .has-continuation {
    display: block !important;
  }
}
```

# Prologue

---
name: schedule

# Schedule

## Today

Dataframes, `dplyr`
- Verbs
- Merge
- Clean

## Upcoming
- Data Visualisation, `ggplot2`
- Loops and Functions
- Simulation, `furrr`
- Empirical Approaches

---

layout: true

# Introduction

---

## Loading/Filtering Data

Usually when using dataframes, we need to get our hands dirty. We will evaluate our options with `base` functions before using functions from `dplyr`.

--

It may well be the case that there is far more data available than we will need. Three options; 

- Cherry pick variables from the source, 
- Trim variables from the file, 
- Load entire file on R and trim down.

---

```{r}
p_load(gapminder)
head(gapminder)
```

Let's see some common `dplyr` functions using the gapminder dataframe. 

---

To generate a variable in your dataframe use `%>% mutate()`
```{r eval=FALSE, include=TRUE}
data_lnGDP <- gapminder %>% mutate( GDP = pop*gdpPercap,
                                    lnGDP = log(GDP)) 
```

To filter out particular rows from your dataframe use `%>% filter()`
```{r eval=FALSE, include=TRUE}
EurAsia <- data_lnGDP %>% filter(continent %in% c("Asia", "Europe"))
# How many countries did I remove?
length(unique(gapminder$country)) - length(unique(EurAsia$country))
```

To summarize by groups, combine `%>% group_by()` and `%>% summarize`<br>
`desc` places `arrange` variables in descending order
```{r eval=FALSE}
sum_EurAsia <- EurAsia %>% group_by(country) %>% summarise(
    avg_pop = mean(pop),avg_gdp = mean(GDP)) %>% arrange(desc(avg_gdp))
sum_EurAsia
```



---

layout: true

# Merging

---

## Binding

#### Binding vectors

Consider `rbind` and `cbind`: they treat the inputs as either rows or columns, and then binds them together.

```{r bind_vector_data, include=F}
name <- c("Pam", "George", "Sandy")
favorite <- c("Glazed Yams", "Leeks", "Daffodils")

# What are the dimensions of these?
rbind(name, favorite)
cbind(name, favorite)
```

```{r bind_vector_data1, eval=F}
name <- c("Pam", "George", "Sandy")
favorite <- c("Glazed Yams", "Leeks", "Daffodils")

# What are the dimensions of these?
rbind(name, favorite)
cbind(name, favorite)
```

--

`rbind` yields a 2x3 

--

`cbind` yields a 3x2

---

## Binding data frames

You can also use `rbind` and `cbind` to bind data frames.
```{r bind_df_data}
# Create some data frames for us to work with
name_fav <- cbind(name, favorite)
name_work <- cbind(name, work = c("Bus Driver", NA, "Shopkeeper"))
name_fav
name_work
```

---

## Binding data frames

`cbind` treats the objects as columns, so they're put side-by-side:
$$\begin{bmatrix}
  A, B \\
\end{bmatrix}$$

```{r cbind_data_frames}
cbind(name_fav, name_work)
```

---

## Binding data frames

`rbind` treats the objects as rows, so they're stacked:
$$\begin{bmatrix}
  A \\
  B \\
\end{bmatrix}$$

```{r rbind_data_frames}
rbind(name_fav, name_work) #notice how rbind doesn't care about column names
```

---

## Binding data frames

`dplyr` has very similar functions `bind_rows` and `bind_cols`. They work best with tibbles, so we'll go ahead and create tibble versions of our data.

```{r as_tibbles}
name_fav_tib <- as_tibble(name_fav)
name_work_tib <- as_tibble(name_work)
```

--

.pull-left[
```{r show, echo=FALSE}
name_fav_tib
```
]

.pull-right[
```{r show2, echo=FALSE}
name_work_tib
```
]

---

## Binding data frames
.pull-left[
```{r bind_cols}
bind_cols(name_fav_tib, name_work_tib)
```
]

.pull-right[
```{r bind_rows}
bind_rows(name_fav_tib, name_work_tib)
```
]

---

## Set Operations

The `dplyr` set operation functions are `union`, `intersect`, and `setdiff`. These set operations treat observations (rows) as if they were set elements.

```{r set_op_data}
table_1 <- tribble(
  ~"name", ~"favorites",
  #------|--------
  "Pam", "Glazed Yams",
  "George", "Leeks",
  "Sandy", "Daffodils"
)

table_2 <- tribble(
  ~"name", ~"favorites",
  #------|--------
  "Pam", "Glazed Yams",
  "Gus", "Fish Tacos"
)
```

---

## Set Operations

Create tibbles using an easier to read row-by-row layout. This is useful for small tables of data where readability is important

.pull-left[
```{r tribble1}
table_1
```
]

.pull-right[
```{r tribble2}
table_2
```
]

---

## Set Operations

`union` will give you all the observations (rows) that appear in either or both tables. This is similar to `bind_rows`, but `union` will remove duplicates.

```{r set_union}
union(table_1, table_2)
```

---

## Set Operations

`intersect` will give you only the observations that appear both in `table_1` and in `table_2`: in the intersection of the two tables.

```{r set_intersect}
intersect(table_1, table_2)
```

---

## Set Operations

`setdiff(table_1, table_2)` gives you all the observations in table_1 that are not in table_2.

```{r set_setdiff}
setdiff(table_1, table_2)
```

---

## Mutating joins

Mutating joins take the first table and add columns from the second table. There are 3 mutating joins: `left_join`, `inner_join`, and `full_join`.

```{r mutating_join_data}
# We'll create 2 new data frames to learn mutating joins:

favorites <- tribble(
  ~"name", ~"fav",
  #------|--------
  "Pam", "Glazed Yams",
  "George", "Leeks",
  "Sandy", "Daffodils"
)

jobs <- tribble(
  ~"name", ~"work",
  #------|--------
  "Pam", "Bus Driver",
  "Gus", "Bartender",
  "Sandy", "Shopkeeper"
)
```

---

## Mutating joins

`left_join(x, y)` takes x and adds the columns of y where the **key** matches. The **key** is a variable that shows up in both tables and you'll specify it with `by = "key_variable"`.

.pull-left[
```{r left_join}
left_join(favorites, jobs, by = "name")
```
]

.pull-right[
```{r Example, eval=F}
# What will be the output?
left_join(jobs, favorites, by = "name")
```
]

---

## Mutating joins

`left_join(x, y)` takes x and adds the columns of y where the **key** matches. The **key** is a variable that shows up in both tables and you'll specify it with `by = "key_variable"`.

.pull-left[
```{r left_join_1}
left_join(favorites, jobs, by = "name")
```
]

.pull-right[
```{r Example_1}
# What will be the output?
left_join(jobs, favorites, by = "name")
```
]

---

## Mutating joins

`inner_join(x, y)` takes the **intersect** of the key variable and adds columns from both tables.
```{r inner_join}
inner_join(favorites, jobs, by = "name")
```

---

## Mutating joins

`full_join(x, y)` takes the **union** of the key variable and adds columns from both tables.
```{r full_join}
full_join(favorites, jobs, by = "name")
```

---

## Filtering joins

Unlike mutating joins, filtering joins will only preserve data from the first table. The observations that are kept depends on the second table. dplyr has 2 types of filtering joins: `semi_join` and `anti_join`.

`semi_join(x, y)` keeps all rows in x where the key matches in y.

```{r semi_join}
semi_join(favorites, jobs, by = "name")
```

---

## Filtering joins

`anti_join(x, y)` keeps rows in x as long as the key **doesn't** have a match in y.
```{r anti_join}
anti_join(favorites, jobs, by = "name")
```

---

## Pivoting

`pivot_wider()` and `pivot_longer()` aren't two-table topics, but they are useful data manipulation tools in the tidyverse.

```{r pivot_data}
prefs <- tribble(
  ~"name", ~"preference", ~"item",
  #|-----|--------------|--------|
  "Pam", "loves", "Glazed Yams",
  "Pam", "likes", "Daffodils",
  "Pam", "hates", "Horseradish",
  "George", "loves", "Leeks",
  "George", "likes", "Hazelnuts",
  "George", "hates", "Dandelions"
)
```

Take a look at the data. There are 2 people (Pam and George). Each person has one "love", one "like", and one "hate" item.

---

## Pivoting

Suppose instead we wanted our data in a different format. What if we had 4 columns instead of 3: `name`, the thing that person `loves`, the thing that person `likes`, and the thing that person `hates`. We'd only need 2 rows (Pam and George).

We want our data to go from having 3 columns to having 4, so we know we can use `tidyr::pivot_wider`.

```{r pivot_wider}
pivot_wider(prefs, names_from = preference, values_from = item)
```

```{r pivot_wider_1, include=FALSE}
prefs_wide <- pivot_wider(prefs, names_from = preference, values_from = item)
```

---

## Pivoting

Now suppose we want to reverse that operation!
We'll start with `prefs_wide` and pivot in to get `prefs` again.

`pivot_longer()` has these arguments: 

 - **cols**: columns to pivot into the longer format. For us, that will be the columns `loves`, `likes`, and `hates`. We can also say columns 2 through 4: `cols = 2:4`.
 - **names_to**: A string. What we should call the new column that holds those old column names: `loves`, `likes`, `hates`: `names_to = "preferences"`
 - **values_to**: A string. what we should call the values that are now being pivoted in? `Glazed Yams`, `Daffodils`, etc. So we want `values_to = "items"`
 
 
---

## Pivoting

```{r pivot_longer}
prefs_wide %>% pivot_longer(cols = 2:4, names_to = "preferences", values_to = "items")
```




---
layout: true

# Cleaning

---

Outliers may also be present in data. In macroeconomics, one may be trying to assess mean and variance of real gross domestic product in the United States.

```{r DataLoad, fig.width = 7, fig.height = 4, echo=FALSE}
#Data Load
library(readxl)
GDPC1 <- read_excel("GDPC1.xls", skip = 10)

fGDP <- diff(GDPC1$GDPC1)
fGDP <- c(NA, fGDP)
GDPC1 <- cbind(GDPC1, fGDP)

#GDP_plot <- ggplot(GDPC1[2:245,]) +
GDP_plot <- ggplot(GDPC1) +
  geom_line(mapping=aes(x=observation_date, y=fGDP), color="blue", linetype = 1, size = 1) +
geom_smooth(mapping=aes(x=observation_date, y=fGDP))+
  theme(plot.title = element_text(hjust = 0.5))+
  theme_pander( base_size = 12) + xlab("Year") + ylab("Change in Real Gross Domestic Product, USD BN") +
  ggtitle("Real Gross Domestic Product, First-Difference, Quarterly")
GDP_plot
```

---
```{r clean, echo=FALSE}
GDP_plot <- ggplot(GDPC1[2:245,]) +
  geom_line(mapping=aes(x=observation_date, y=fGDP), color="blue", linetype = 1, size = 1) +
geom_smooth(mapping=aes(x=observation_date, y=fGDP))+
  theme(plot.title = element_text(hjust = 0.5))+
  theme_pander( base_size = 12) + xlab("Year") + ylab("Change in Real Gross Domestic Product, USD BN") +
  ggtitle("Real Gross Domestic Product, First-Difference, Quarterly")
GDP_plot
```

---

Consider the 1.5 IQR rule of thumb. This is used to identify mild outliers. For extreme outliers only, shift to a 3 IQR.

+ IQR = 75th Percentile Value - 25th Percentile Value 
+ Lower Outlier Boundary = 25th - 1.5*IQR
+ Upper Outlier Boundary = 75th + 1.5*IQR

```{r IQR, eval=F}
#Extreme Outliers identified and cleaned
Phase1 <- summary(data$columnX)
OutLower <- Phase1[2]-3*(Phase1[5]-Phase1[2])
OutHigher <- Phase1[5]+3*(Phase1[5]-Phase1[2])

house_w <- filter(house, columnX > OutLower)
house_w <- filter(house, columnX < OutHigher)
```

See example of extreme outlier cleaning in Davies, R., & T., Jeppesen, 2015 *"Export mode, firm heterogeneity, and source country characteristics*, Review of World Economics, Vol. 151(2), pp 169-195.

---

## Resources

* [Datacamp](https://learn.datacamp.com/courses/joining-data-with-dplyr-in-r): Joining data with dplyr 
* [RStudio dplyr Cheat Sheet](https://rstudio.com/wp-content/uploads/2015/02/data-wrangling-cheatsheet.pdf)

---

exclude: true

```{R generate pdfs, include = F, eval = F}
#remotes::install_github('rstudio/pagedown')
library(pagedown)
pagedown::chrome_print("Tutorial-Slides-4.html", output = "Tutorial-Slides-4.pdf")
#pagedown::chrome_print("Tutorial Slides 1-nopause.html", output = "01-research-r-nopause.pdf")
```